Fine-scale Data Reveals Neighborhood Inequalities In Effects of Heat on Expressed Mood

### 2021-08-19 doc/pnas/PNAS-template-main.tex
pnas submission

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{breakcites}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
% \usepackage[
% style=numeric,
% citestyle=numeric-comp,
% backend=biber,
% doi=true,
% natbib=true,
% sorting=none
% ]{biblatex}
\usepackage{scalerel,xparse}

% To be able to use emojis
\NewDocumentCommand\emojismile{}{
 \scalerel*{
 \includegraphics{smiling-face-with-smiling-eyes_1f60a.png}
 }{X}
}

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Data Reveals Neighborhood Inequalities In Effects of Heat on Expressed Mood}

\author{}
% \author[1,*]{Matthew Cooper}
% \author[2]{Jeremiah Osborne-Gowey}
% \author[3]{Zheng Liu}
% \author[4]{Portia Adade Williams}
% \author[5]{Jie Liu}
% \author[6]{Aaron Schwartz}
% \author[7]{Patrick Baylis} 
% 
% \affil[1]{T.H. Chan School of Public Health, Harvard University, U.S.A.}
% \affil[2]{Environmental Studies Program, University of Colorado Boulder, U.S.A.}
% \affil[3]{Department of Geographical Sciences, University of Maryland College Park, U.S.A.}
% \affil[4]{CSIR-Science and Technology Policy Research Institute, Ghana}
% \affil[5]{School of Business, East China University of Science and Technology, China} 
% \affil[6]{Department of Ecology \& Evolutionary Biology, University of Colorado Boulder, U.S.A.}
% \affil[7]{Vancouver School of Economics, University of British Columbia, Canada}
% \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

\begin{abstract}
%Limit 150 words:

%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health, and previous work has found strong associations between heat and mental well-being. However, these studies typically use data reported at the city or county level, and, to date, have found no difference in vulnerability based on income or race.  We therefore use expressed mood in a quarter of a billion geolocated tweets as a proxy for mental health status available at fine spatial and temporal scales, and find stark differences in the effects of heat depending on neighborhood characteristics. Here we show that increased temperatures worsen expressed mood in all areas, but that this effect is much stronger in poor and Black neighborhoods. We also find that the effect of heat on expressed mood is greatest in the early morning, supporting the hypothesis that heat affects mental health through a sleep quality pathway. This paper presents the first evidence of neighborhood heterogeneity in the vulnerability of mental health to heat.
\end{abstract}
\begin{document}

\raggedbottom
\maketitle
\thispagestyle{empty}

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being. A large body of research has explored the ongoing impacts of climate change on agricultural outcomes, economic growth, and physical health \cite{pachauri2014climate}. However, academic and medical researchers have highlighted the relative paucity of empirical studies on the impacts of climate change on mental health \cite{Berry2018Apr}. While mental health is under-studied, it accounts for up 13\% of the global burden of disease, representing a major share of total human suffering \cite{Collins2011Jul}, leading to calls for more research into potential linkages between climate change and mental health \cite{Berry2018Apr, Collins2011Jul}. Researchers have begun to answer these calls, finding associations between climate disasters and post-traumatic stress \cite{Schwartz2017Aug}, as well as linkages between higher temperatures and a variety of mental health outcomes \cite{baylis_weather_2018, Mullins2019Dec, Obradovich2018Oct}. Nevertheless, with average global temperatures projected to increase by 1.5\textdegree C within a decade \cite{allen2019technical}, more research is needed into the impacts of increased heat on mental health, especially for identifying the most vulnerable communities.

%Theoretical: frameworks and pathways
Researchers have proposed a number pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr}. Work on biological mechanisms emphasizes the mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul}. Other studies point to linkages between increased temperatures, overall physical health and increases in injuries which can in turn affect mental health \cite{Berry2007}. Additionally, exposure to higher temperatures can affect productivity and income \cite{Burke2015Nov}, leading to second-order impacts on mental health outcomes.

@@

The effects of heat exposure on mental health are by now relatively well established, yet results from research conducted at national scales indicates surprisingly little heterogeneity in impacts, with consequences for our understanding of groups that are most vulnerable as well as potential adaptations to future heat stress. For example, a large study of temperature and suicide in the USA and Mexico found "no significant difference in suicide response to temperature between rich and poor municipalities or counties" \cite{Burke2018Aug}. Similarly, a national study found no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status across the USA \cite{Mullins2019Dec}. A major challenge for these studies, however, is their reliance on data that is typically aggregated to the municipality or county level. As a consequence, these studies are often restricted to between-county metrics of vulnerability and cannot take into account neighborhood effects.

These findings of uniform vulnerability to heat are surprising given that minority groups and low-income people are more likely to be exposed to undesirable temperatures and less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality air-conditioning \cite{Samuelson2020Jun}. Many of these factors that expose low-income people to heat may be exacerbated and compounded for racial minorities. For example, Black Americans face discrimination in employment \cite{Kang2016, Penner2008} and housing access \cite{Desmond2015, Akbar2019}, forcing them into jobs and homes that may expose them to more heat. Additionally, policymakers have historically been less likely to take measures to address environmental hazards in Black neighborhoods \cite{Banzhaf2019}. Some studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical and mental health \cite{Belanger2015Mar, Uejio2011Mar}, as well as for vulnerability to other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}. Thus, there are theoretical reasons to expect heterogeneity in the impact of heat on mental health outcomes for different populations, particularly where socioeconomic, racial and other social inequities intersect, although there is little evidence of this to date.

@@

\section*{Effect of Heat on Sentiment}
We examine the relationship between the mood expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. We find higher temperatures are associated with worsened mood (Fig. \ref{fig:wbgt}) after controlling for a variety of fixed effects across all tweets. Mood is highest at 5\textdegree C WBGT (associated in our data set with a median dry bulb temperature on 12\textdegree C/54\textdegree F) with the largest declines between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Mood also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. Dis-aggregating our analysis by metropolitan area and using alternative methods to estimate expressed mood all yield similar results (see Supplement).

\begin{figure}[H]
 \centering
 \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
 \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and expressed mood in tweets. As temperatures increase above 5\textdegree C WBGT, mood rapidly declines.}
 \label{fig:wbgt}
\end{figure}

We find that mean neighborhood income strongly moderates the relationship between temperature and mood (Fig. \ref{fig:hetero}), with large differences in mood between the poorest (5th percentile) and wealthiest (95th percentile) neighborhoods. As temperatures increase to a modest 20\textdegree C WBGT (29\textdegree C/84\textdegree F), mood increases in the wealthiest neighborhoods but decreases in the median and lowest income neighborhoods. The wealthiest neighborhoods do not see decreases in mood until temperatures exceed 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point mood decreases relatively evenly regardless of neighborhood income percentile.

We also explored how neighborhood racial characteristics affect the relationship between temperature and mood, and we find the effects of heat are felt disproportionately in neighborhoods that are majority Black (Fig. \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the mood of tweets in majority Black neighborhoods decrease four times as much as the mood of people in other neighborhoods. Additionally, at mild to warm temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower mood than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

\begin{figure}[H]
\centering
 \includegraphics[width=\linewidth]{../res/wbgt_combined.png}
 \caption{Effect of changes in WBGT on expressed mood as moderated by neighborhood income (a) and race (b).}
\label{fig:hetero}
\end{figure}

Finally, given that both low-income and majority black neighborhoods are more vulnerable to heat waves, we model the combined effects of race and income together to explore the intersection of these two types of vulnerability. Specifically, we model how the relationship between temperature and mood is moderated by the interaction between a neighborhood's per-capita income and the percentage of a neighborhood's inhabitants that are black. Additionally, we compare AICs for models with no term for vulnerability, a race term for vulnerability, an income term for vulnerability, and both race and income terms for vulnerability.

\begin{minipage}{.65\textwidth}
\centering
\includegraphics[width=\textwidth]{../res/wbgt-income-race.png}
\captionof{figure}{Relationship between WBGT and mood from text in tweets, moderated by the interaction of race and income.}
\label{fig:race-income}
\end{minipage}\hfill
\begin{minipage}{.35\textwidth}
\centering
\begin{tabular}{| l | l |}
\hline
Terms & AIC \\
\hline
Neither & 277,270,561 \\
Income & 277,182,239 \\
Race & 277,143,579 \\
Both & 277,108,676 \\
\hline
\end{tabular}
\captionof{table}{Akaike information criterion (AIC) of models with interaction terms for race, income, both race and income, as well as neither race nor income. The model with the best fit by AIC includes both race and income interaction terms.}
\label{tab:aics}
\end{minipage}

Graphing the combined effects of race and income, we find that race has a larger impact on vulnerability than income (see Fig. \ref{fig:race-income}). A Black neighborhood at any income level will be much more impacted by heat than a poor neighborhood with few black inhabitants. Additionally, neighborhoods that are largely black are also more affected by cold temperatures, although there is no clear effect of income moderating the relationship between cold temperatures and mood. Using AIC as an indicator of how different terms affect model quality (see Tab. \ref{tab:aics}), we find that a model including both terms performs the best, indicating that both race and income matter for the impact of heat on mood. However, including only a race term improves the model more than including only an income term, further supporting the conclusion that race is more important than income for explaining the vulnerability of mood to heat.

\section*{Comparison With Other Events}
We compared the expressed mood from exposure to temperatures to two other events associated with impacts on mood and other mental health outcomes: the weekly change from Saturday to Monday, as well as the impact of a major natural disaster (see Fig. \ref{fig:compare}). For this comparison, we define heat impact as the changes in mood associated with temperatures increasing from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F); we define weekly changes as the difference between the mean Saturday mood and the mean Monday mood, calculated across the entire data set; and we define the impact of Hurricane Sandy as the change in mood from the week before the hurricane made landfall to the week after, calculated for only impacted counties. Both of these events are associated with mental health impacts: the decrease in expressed mood on Twitter from Saturdays to Mondays mirrors mirrors the weekly dynamics of suicide rates \cite{CDC2021}, while Hurricane Sandy was associated with mental health effects on victims including increased anxiety, PTSD, and depression \cite{Schwartz2017Aug, Lieberman-Cribbin2017}.

\begin{figure}[H]
 \centering
 \includegraphics[width=0.66\linewidth]{../res/comparison_plot.png}
 \caption{Decline in mood as temperature changes from mild (5\textdegree C WBGT) high (25\textdegree C WBGT) across different neighborhood types, compared with the impacts of the change from Saturday to Monday, as well as the change associated with Hurricane Sandy.}
 \label{fig:compare}
\end{figure}

We find that the decline in mood in wealthier and majority white neighborhoods from exposure to increasing temperatures is less than the average weekly change in mood from Saturday to Monday (see Fig. \ref{fig:compare}). Conversely, the effects of exposure to increasing temperatures in poorer and majority Black neighborhoods are much larger than the average weekly change in mood. Additionally, the impacts of exposure to these changes in temperature for the more vulnerable neighborhoods are close in severity to the impact on mood from experiencing a major hurricane.

\section*{Effect of Heat by Time of Day}
In addition to providing data with high spatial resolution, Twitter data also comes with very high temporal resolution as each tweet is time-stamped. Thus, we were able to examine the impact of heat on mood over the course of the day (Fig. \ref{fig:ts-wbgt}). We found that heat is associated with improved mood for a brief period in the late morning, and then has a negative effect on mood during the latter half of the day, with a consistent effect from noon until 9pm. This effect weakens in the early evening through midnight, and then increases substantially throughout the night. Heat has the greatest effect on expressed mood at 6am, an effect over twice as large as during the day.

\begin{figure}[H]
 \centering
 \includegraphics[width=0.5\linewidth]{../res/ts_heat.png}
 \caption{Effects of rising temperatures on mood by hour of the day, with a 95\% confidence interval. The value shown is the predicted change in VADER score for a 10\textdegree C WBGT increase in temperatures.}
 \label{fig:ts-wbgt}
\end{figure}

\section*{Discussion}
% Compare weekly changes in mood to weekly changes in suicides
We found that the change from an optimum temperature to a higher temperature is associated with an overall decrease in mood similar in magnitude to the degree of change in mood from Monday to Saturday, a weekly change associated with an increase in suicides of 23\% over the course of the study period \cite{CDC2021}, and that the same level of heat in a poor or Black neighborhood is associated with a change in mood nearly matching the impact of a major Hurricane, an event associated with suicides and PTSD \cite{Schwartz2017Aug, Lieberman-Cribbin2017}.  Thus, while research on linkages between expressed mood and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of expressed mood and mental health at specific time scales and following certain events.  This suggests that the changes in expressed mood we observe at neighborhood scales is indicative of other mental health outcomes such as suicides and hospitalizations.

% Discuss the NOVELTY: quote Burke and Mullins, refute their interpretations, people have said XX, but we have shown ~XX.
% Variations in vulnerability also indicates ADAPTATION
Previous work conducted with county-scale data found no heterogeneity in mental health vulnerability by income. Such analyses have led to conclusions that the mental health effects of climate change will be uniform, least in developed countries, and that adaptation with technologies like air-conditioning will not mitigate the mental health effects of climate change. For example, Mullins et al. conclude that ``individuals have not been able to successfully reduce the negative effects of higher temperatures on mental health" \cite{Mullins2019Dec}. Contrary to these analyses, we found large heterogeneity in vulnerability by neighborhood racial composition and per-capita income. This suggests that the mental health effects of climate change will not be uniform and, like other impacts, will fall disproportionately on the vulnerable and marginalized. More positively, it does suggest that adaptation is possible, and, for communities with infrastructure and working conditions similar to those of higher-income Americans, the effects of high temperatures on mental health may be addressable.

@@

% Connect to bigger literature on vulnerability
This study shows the importance of examining heterogeneities in vulnerability in analyses of climate change impacts.  Additionally, it shows how measuring outcomes at fine spatial scales necessary for understanding how factors like race and income matter for vulnerability \cite{Tong2021}. Even within a relatively wealthy nation, we find that high temperatures affect certain populations to the same degree as a major hurricane. Thus, it is likely that heat waves have a much greater mental health impact throughout the developing world, where temperatures are expected to be much higher \cite{Raymond2020May}, heatwaves are already under-counted \cite{Harrington2020Sep}, and adaptive technologies like air conditioning are more scarce \cite{Biardeau2020Jan}.

In addition to exploring heterogeneities in vulnerability, using twitter data lets us explore how the effect of heat varies by time of date and infer possible important pathways in the heat-mental health relationship. Recent research has suggested that the impacts of heat on sleep quality may play a large role in the observed mental health effects of heat \cite{Obradovich2017May, Mullins2019Dec}, and our temporal analysis largely supports this hypothesis.  We found much stronger effects of heat on expressed sentiment in the early morning, adding weight to the sleep quality pathway linking heat and mental health that other authors have found.  This suggests that efforts to improve mental health during heat waves may have the largest impact by focusing on providing electricity and air conditioning at night.

% Caveats
% Mood is a fuzzy indicator but we have BIG DATA
While these findings are robust to different model specifications and metrics of mood, there are important considerations in this analysis.  One caveat associated with this data is that we were only able to locate the tweets within the census block from which the tweet was sent, and in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are found that census block. For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day. These issue may in fact lead us to under-estimate the true effect of heat on mental health for poor and Black people. Another issue with using Twitter data is that, while Twitter is used by more than one in five Americans and at similar rates across racial groups, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}. Nevertheless, we found a large volume of tweets across all neighborhood types.

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{Fine-scale Data Reveals Neighborhood Inequalities In Effects of Heat on Expressed Mood}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Matthew Cooper}
\author[b]{Jeremiah Osborne-Gowey}
\author[c]{Zheng Liu}
\author[d]{Portia Adade Williams}
\author[e]{Jie Liu}
\author[f]{Aaron Schwartz}
\author[g]{Patrick Baylis}

\affil[a]{T.H. Chan School of Public Health, Harvard University, 677 Huntington Avenue, Boston, MA 02115, USA}
\affil[b]{Environmental Studies Program, University of Colorado Boulder, 4001 Discoverty Drive, Boulder, CO 80303, USA}
\affil[c]{Department of Geographical Sciences, University of Maryland College Park, 7251 Preinkert Drive, College Park, MD 20742, USA}
\affil[d]{CSIR-Science and Technology Policy Research Institute, Agostinho Neto Road, Accra, Ghana}
\affil[e]{School of Business, East China University of Science and Technology, Meilonglu Street 130, Shanghai 2000237, China} 
\affil[f]{Department of Ecology \& Evolutionary Biology, University of Colorado Boulder, 1900 Pleasant Street, Boulder, CO 80309, USA}
\affil[g]{Vancouver School of Economics, University of British Columbia, 6000 Iona Drive
Vancouver, BC, V6T 1L4, Canada}

% Please give the surname of the lead author for the running footer
\leadauthor{Cooper}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Higher temperatures are known affect many mental health outcomes, including suicides, hospitalizations, and expressed mood in social media.  However, work to date has found that heat affects mental health uniformly across all populations.  This is surprising, given that people vary greatly in their exposure to heat based on their working, commuting, and housing conditions.  We draw on fine-scale twitter data to show that temperature does not affect the mental health of all populations the same, and that people in poor and Black neighbors are much more affected by the heat.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{M.C., J.O.-G., A.S. and P.B. designed the research and modeling strategy; A.S. provided data; M.C and Z.L. prepared data; M.C. and J.O.-G. analyzed data; and M.C., J.O.-G., P.A.W., and J.L. wrote the paper.}
\authordeclaration{We have no competing interests to decalre.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: mcooper@hsph.harvard.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Heat $|$ Mental Health $|$ Inequity $|$ Big Data}

\begin{abstract}
  %Up to 250 words
Higher temperatures associated with climate change are expected to have major impacts on human mental health, and previous work has found strong associations between heat and mental well-being. However, these previous studies typically used data reported at the city or county level, and, to date, have found no difference in vulnerability based on income or race.  We therefore use expressed mood in a quarter of a billion geolocated tweets as a proxy for mental health status available at fine spatial and temporal scales, and pair each tweet with data on local temperature and neighborhood demographics.  We show that heat affects expressed mood in all areas, but find that this effect is much stronger in poor and Black neighborhoods.  We also find that the effect of heat on expressed mood is greatest in the early morning, supporting the hypothesis that heat affects mental health through a sleep quality pathway. This paper presents the first evidence of neighborhood heterogeneity in the vulnerability of mental health to heat.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

%Overview
\dropcap{C}limate change is impacting many aspects of human well-being. A large body of research has explored the ongoing impacts of climate change on agricultural outcomes, economic growth, and physical health \cite{pachauri2014climate}. However, academic and medical researchers have highlighted the relative paucity of empirical studies on the impacts of climate change on mental health \cite{Berry2018Apr, Berrang-Ford2021}. While mental health is under-studied, it accounts for up 13\% of the global burden of disease, representing a major share of total human suffering \cite{Collins2011Jul}, leading to calls for more research into potential linkages between climate change and mental health \cite{Berry2018Apr, Collins2011Jul}. Researchers have begun to answer these calls, finding associations between climate disasters and post-traumatic stress \cite{Schwartz2017Aug}, as well as linkages between higher temperatures and a variety of mental health outcomes \cite{baylis_weather_2018, Mullins2019Dec, Obradovich2018Oct}. Nevertheless, with average global temperatures projected to increase by 1.5\textdegree C within a decade \cite{allen2019technical}, more research is needed into the impacts of increased heat on mental health, especially for identifying the most vulnerable communities.

%Theoretical: frameworks and pathways
Researchers have proposed a number pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr}. Work on biological mechanisms emphasizes the mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul}. Other studies point to worsened mental health as a second-order effect of the direct impacts of increased temperature on overall physical health and risk of injury \cite{Berry2007}, as well as productivity and income \cite{Burke2015Nov}.

@@

The effects of heat exposure on mental health are by now relatively well established, yet results from research conducted at national scales indicates surprisingly little heterogeneity in impacts, with consequences for our understanding of groups that are most vulnerable as well as potential adaptations to future heat stress. For example, a large study of temperature and suicide in the USA and Mexico found ``no significant difference in suicide response to temperature between rich and poor municipalities or counties'' \cite{Burke2018Aug}. Similarly, a national study found no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status across the USA \cite{Mullins2019Dec}. A major challenge for these studies, however, is their reliance on data that is typically aggregated to the municipality or county level. As a consequence, these studies are often restricted to between-county metrics of vulnerability and cannot take into account heterogeneity at more local scales.

These findings of uniform vulnerability to heat are surprising given that minority groups and low-income people are more likely to be exposed to undesirable temperatures and less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}; more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}; and more likely to live in housing that is less insulated with poorer quality air-conditioning \cite{Samuelson2020Jun}. Many of these factors that expose low-income people to heat may be exacerbated and compounded for racial minorities. For example, Black Americans face discrimination in employment \cite{Kang2016, Penner2008} and housing access \cite{Desmond2015, Akbar2019}, forcing them into jobs and homes that may expose them to more heat. Additionally, policymakers have historically been less likely to take measures to address environmental hazards in Black neighborhoods \cite{Banzhaf2019}. Some studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health \cite{Belanger2015Mar, Uejio2011Mar}, as well as for vulnerability to other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}. Thus, there are theoretical reasons to expect heterogeneity in the impact of heat on mental health outcomes for different populations, particularly where socioeconomic, racial and other social inequities intersect, although there is little evidence of this to date.

@@

\section*{Results}
\subsection*{Effect of Heat on Sentiment}
We examine the relationship between the mood expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. We find higher temperatures are associated with worsened mood (Fig. \ref{fig:wbgt}) after controlling for a variety of fixed effects across all tweets. Mood is highest at 5\textdegree C WBGT (associated in our data set with a median dry bulb temperature on 12\textdegree C/54\textdegree F) with the largest declines between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Mood also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. Dis-aggregating our analysis by metropolitan area and using alternative methods to estimate expressed mood all yield similar results (see Supplement).

\begin{figure}[H]
 \centering
 \includegraphics[width=\linewidth]{../../res/wbgt.png}
 \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and expressed mood in tweets. As temperatures increase above 5\textdegree C WBGT, mood rapidly declines.}
 \label{fig:wbgt}
\end{figure}

We find that mean neighborhood income strongly moderates the relationship between temperature and mood (Figs. \ref{fig:income}), with large differences in mood between the poorest (5th percentile) and wealthiest (95th percentile) neighborhoods. As temperatures increase to a modest 20\textdegree C WBGT (29\textdegree C/84\textdegree F), mood increases in the wealthiest neighborhoods but decreases in the median and lowest income neighborhoods. The wealthiest neighborhoods do not see decreases in mood until temperatures exceed 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point mood decreases relatively evenly regardless of neighborhood income percentile.

We also explored how neighborhood racial characteristics affect the relationship between temperature and mood, and we find the effects of heat are felt disproportionately in neighborhoods that are majority Black (Fig. \ref{fig:race}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the mood of tweets in majority Black neighborhoods decrease four times as much as the mood of people in other neighborhoods. Additionally, at mild to warm temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower mood than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

\begin{figure}[H]
\centering
 \includegraphics[width=0.8\linewidth]{../../res/wbgt-income.png}
 \caption{Effect of changes in WBGT on expressed mood as moderated by neighborhood income per capita.}
\label{fig:income}
\end{figure}

\begin{figure}[H]
\centering
 \includegraphics[width=0.8\linewidth]{../../res/wbgt-race_q.png}
 \caption{Effect of changes in WBGT on expressed mood as moderated by neighborhood majority race.}
\label{fig:race}
\end{figure}

\subsection*{Comparison With Other Events}
We compared the change in expressed mood from heat exposure to two other events associated with impacts on mood and other mental health outcomes: the weekly change from Saturday to Monday, as well as the impact of a major natural disaster (see Fig. \ref{fig:compare}). For this comparison, we define heat impact as the changes in mood associated with temperatures increasing from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F); we define weekly changes as the difference between the mean Saturday mood and the mean Monday mood, calculated across the entire data set; and we define the impact of Hurricane Sandy as the change in mood from the week before the hurricane made landfall to the week after, calculated for only impacted counties. Both of these events are associated with mental health impacts: the decrease in expressed mood on Twitter from Saturdays to Mondays mirrors mirrors the weekly dynamics of suicide rates \cite{CDC2021}, while Hurricane Sandy was associated with mental health effects on victims including increased anxiety, PTSD, and depression \cite{Schwartz2017Aug, Lieberman-Cribbin2017}.

\begin{figure}[H]
 \centering
 \includegraphics[width=0.9\linewidth]{../../res/comparison_plot.png}
 \caption{Decline in mood as temperature changes from mild (5\textdegree C WBGT) to high (25\textdegree C WBGT) across different neighborhood types, compared with the impacts of the change from Saturday to Monday, as well as the change associated with Hurricane Sandy.}
 \label{fig:compare}
\end{figure}

We find that the decline in mood in wealthier and majority white neighborhoods from exposure to increasing temperatures is less than the average weekly change in mood from Saturday to Monday (see Fig. \ref{fig:compare}). Conversely, the effects of exposure to increasing temperatures in poorer and majority Black neighborhoods are much larger than the average weekly change in mood. Additionally, the impacts of exposure to these changes in temperature for the more vulnerable neighborhoods are close in severity to the average impact on mood of experiencing a major hurricane.

\subsection*{Effect of Heat by Time of Day}
In addition to providing data with high spatial resolution, Twitter data also comes with very high temporal resolution as each tweet is time-stamped. Thus, we were able to examine the impact of heat on mood over the course of the day (Fig. \ref{fig:ts-wbgt}). We found that heat is associated with improved mood for a brief period in the late morning, and then has a negative effect on mood during the latter half of the day, with a consistent effect from noon until 9pm. This effect weakens in the early evening through midnight, and then increases substantially throughout the night. Heat has the greatest effect on expressed mood at 6am, an effect over twice as large as during the day.

\begin{figure}[H]
 \centering
 \includegraphics[width=0.8\linewidth]{../../res/ts_heat.png}
 \caption{Effects of rising temperatures on mood by hour of the day, with a 95\% confidence interval. The value shown is the predicted change in VADER score for a 10\textdegree C WBGT increase in temperatures.}
 \label{fig:ts-wbgt}
\end{figure}

\section*{Discussion}
% Compare weekly changes in mood to weekly changes in suicides
We found that the change from an optimum temperature to a higher temperature is associated with an overall decrease in mood similar in magnitude to the degree of change in mood from Monday to Saturday, a weekly change associated with an increase in suicides of 23\% over the course of the study period \cite{CDC2021}, and that the same level of heat in a poor or Black neighborhood is associated with a change in mood nearly matching the impact of a major Hurricane, an event associated with suicides and PTSD \cite{Schwartz2017Aug, Lieberman-Cribbin2017}.  Thus, while research on linkages between expressed mood and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of expressed mood and mental health at specific time scales and following certain events.  This suggests that the changes in expressed mood we observe at neighborhood scales is indicative of other mental health outcomes such as suicides and hospitalizations that cannot be measured with neighborhood-scale precision.

% Discuss the NOVELTY: quote Burke and Mullins, refute their interpretations, people have said XX, but we have shown ~XX.
% Variations in vulnerability also indicates ADAPTATION
Previous work conducted with county-scale data found no heterogeneity in mental health vulnerability by income. Such analyses have led to conclusions that the mental health effects of climate change will be uniform, at least in developed countries, and that adaptation with technologies like air-conditioning will not mitigate the mental health effects of climate change. For example, Mullins et al. conclude that ``individuals have not been able to successfully reduce the negative effects of higher temperatures on mental health'' \cite{Mullins2019Dec}. Contrary to these analyses, we found large heterogeneity in vulnerability by neighborhood racial composition and per-capita income. This suggests that the mental health effects of climate change will not be uniform and, like other impacts, will fall disproportionately on the vulnerable and marginalized. More positively, it does suggest that adaptation is possible, and, for communities with infrastructure and working conditions similar to those of higher-income Americans, the effects of high temperatures on mental health may be addressable.

@@

% Connect to bigger literature on vulnerability
This study shows the importance of examining heterogeneities in vulnerability in analyses of climate change impacts.  Additionally, it shows how measuring outcomes at fine spatial scales is necessary for understanding how factors like race and income matter for vulnerability \cite{Tong2021}. Even within a relatively wealthy nation, we find that high temperatures affect certain populations to the same degree as a major hurricane. Thus, it is likely that heat waves have a much greater mental health impact throughout the developing world, where temperatures are expected to be much higher \cite{Raymond2020May}, heat waves are already under-counted \cite{Harrington2020Sep}, and adaptive technologies like air conditioning are more scarce \cite{Biardeau2020Jan}.

In addition to exploring heterogeneities in vulnerability, using twitter data lets us explore how the effect of heat varies by time of day and infer possible important pathways in the heat-mental health relationship. Recent research has suggested that the impacts of heat on sleep quality may play a large role in the observed mental health effects of heat \cite{Obradovich2017May, Mullins2019Dec}, and our temporal analysis largely supports this hypothesis.  We found much stronger effects of heat on expressed sentiment in the early morning, adding weight to the sleep quality pathway linking heat and mental health that other authors have found.  This suggests that efforts to improve mental health during heat waves may have the largest impact by focusing on providing electricity and air conditioning at night.

% Caveats
% Mood is a fuzzy indicator but we have BIG DATA
While these findings are robust to different model specifications and metrics of mood, there are important considerations in this analysis.  One caveat associated with this data is that we were only able to locate the tweets within the census block from which the tweet was sent, and in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are found that census block. For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day. These issues may in fact lead us to under-estimate the true effect of heat on mental health for poor and Black people. Another issue with using Twitter data is that, while Twitter is used by more than one in five Americans and at similar rates across racial groups, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}. Nevertheless, we found a large volume of tweets across all neighborhood types.
